---
title: "Policy exploration"
author: "Jongbin Jung"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Policy exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r pkgs, message=FALSE}
# Load necessary packages
library(tidyverse)
library(undi)
```

The purpose of `undi` is to analyze possibilities of (unjustified) disparate
impact, within the context of a policy. The basic requirements of a policy are
(1) an action taken upon each unit (e.g., requiring bail for defendants or
deciding to frisk an individual) and 
(2) a clear outcome of interest (e.g., whether a defendant fails to appear for a
future court date, or whether the individual is carrying an illegal weapon)[^binary].

[^binary]: For now, `undi` only deals with policies that have binary decisions and binary outcomes.

In addition to an action and outcome, analysis of disparate impact requires the
definition of some group (e.g., race or gender) for which we would like to
measure the possibility of disparate impact.


# The setup

For illustration, we generate data from a (fictional) policy. The united nations
of M&M's (yes, the candy) have recently enacted a "Nuts in my backyard" policy,
which aims to reduce the population of potentially fatal peanut M&M's.
As part of this effort, the M&M's police department (MMPD) were given the
authority to stop, question, and slice pedestrian M&M's highly suspected of
containing peanuts.
Many attributes of M&M's, such as size or visible crack in coating, are
indicitive of being of the peanut variety. And while certain colors are indeed
more likely to be peanut M&M's, it is unclear whether the policy has unjustified
disparate impact on certain colors of M&M's.

```{r}
inv_logit <- stats::binomial()$linkinv
logit <- stats::binomial()$linkfun

N <- 3000

set.seed(1)
data <- tibble(id = rep(1:N)) %>%
  mutate(color = factor(sample(c("yellow", "blue", "red", "other"), 
                               N, replace = TRUE, prob = c(.15, .45, .35, .05))), 
    color = fct_relevel(color, "yellow"),
    size = rnorm(N, 0, 1),
    # yellow M&M's are systematically less-likely to have a crack
    p_crack = ifelse(color == "yellow", 0.1, 0.15),
    has_cracks = rbinom(N, 1, p_crack),
    # noise
    e_risk = rnorm(N, 0, .1),
    # risk of being peanut M&M's is a function of size/cracks/color
    risk = size + 2*has_cracks + 
      # blue and red colors are more likely to have peanuts 
      0.5 * (color == "blue") + 0.5 * (color == "red") + e_risk,  
    # MMPD actually slice blue M&M's at a higher rate 
    p_slice = inv_logit(risk + 0.2*(color == "blue")), 
    # MMPD slices 20% of the population, after adjusting for colorism
    slice = p_slice >= quantile(p_slice, .2) , 
    # slice = inv_logit(p_slice) > runif(n()), 
    # The actual outcome is only observed for sliced M&M's 
    has_peanut = ifelse(slice, inv_logit(risk) > runif(n()), FALSE))
```

## Fitting the policy

In this setting, the unit of analysis is each inidividual candy.
The treatment decision is weather or not to slice each candy, 
the outcome (response) is whether the candy contains peanuts, and `color` is the
group of interest.
Note that we only observe weather each M&M's contains peanuts if it is sliced,
i.e.,  the response given control is always 0.

```{r policy fit}
pol <- policy(slice ~ color + size + has_cracks,
              data,
              model = "glm",
              outcome = "has_peanut", 
              resp_ctl = 0)
plot(pol)
```

First, we can compute regular benchmark tests
```{r benchmark test}
compute_bm(pol, base_group = "yellow", minority_groups = c("blue", "red"))
compute_bm(pol, base_group = "yellow", minority_groups = c("blue", "red"), kitchen_sink = TRUE)
```

We can use the estimated risk scores to measure risk-adjusted disparate impact (rad)
```{r rad}
compute_rad(pol, base_group = "yellow", minority_groups = c("blue", "red"))
```
